fGFPCA dynamic prediction (In-sample)

### 2022-11-21 f-GFPCA.Rmd
try in-sample prediction (point)

---
title: "fGFPCA dynamic prediction"
output: html_document
date: "2022-11-03"
---

@@

library(here)
```

## Toy simulation with a very simple function

@@

- Below we generate 100 latent functions and their corresponding binary functions

$$Z_i(t) = a_{i1}sin^2(t)+a_{i2}cos^2(t)+a_{i3}t^2+a_{i4}t$$

$$Y_i(t) \sim Binomial(\frac{exp(Z_i(t))}{1+exp(Z_i(t))})$$

```{r}
## four subjects
ggplot(df_train2 %>% filter(id %in% 15:18))+
  geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1,
             show.legend = F)+
  geom_line(aes(x=argvals, y=Z, col = "Red"), 
            show.legend = F)+
  facet_wrap(~id)+
  labs(title = "Simulated function for the 4 subjects",
       x = "time", y ="")

## all subjects latent functions
ggplot(df_train2)+
  geom_line(aes(x=argvals, y=Z, col = as.factor(id)), 
            show.legend = F)+
  labs(title = "Simulated latent functions for all subjects")

@@

- Below is the predicted latent functions from local GLMM models

```{r}

ggplot(df_pred_latent %>% filter(id %in% 15:18))+
  geom_line(aes(x = argvals, y=Zhat))+
  geom_line(aes(x=argvals, y = Z, col = "red"))+
  geom_point(aes(x=argvals, y = Y, col = "blue"), size = 0.01)+
  facet_wrap(~id)

@@

- Use fastGFPCA to generate more complicated functions
- What happened to the latent gaussian function? Are they just noisy versions of true latent functions? Or it there something wrong with GLMMs? 
- Establish prediction intervals of latent gaussian function
- Write a function to calculate random effects (instead of using predict.fpca.sparse)

>>>>

---
title: "fGFPCA dynamic prediction (In-sample)"
output: html_document
date: "2022-11-03"
---

@@

library(here)
library(tidyverse)
library(ggplot2)
library(fcr)
library(refund)
library(lme4)
set.seed(1025)
```

## Toy simulation

### Simulation set-up

@@

- For a random sample i:

$$\eta_i(t) =f_0(t)+ \xi_{i1}sin(2\pi t)+\xi_{i2}cos(2\pi t)+\xi_{i3}sin(4\pi t)+\xi_{i4}cos(4\pi t)$$

where $\boldsymbol{\xi}_{i} \sim N(0, \boldsymbol{\Gamma})$. In the fPCA framework, $\boldsymbol{\Gamma}$ is a diagonal matrix and diagonal elements corresponds to eigenvalues.

$$Y_i(t) \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))})$$

In this simulation, we set $f_0(t) = 0$, $\boldsymbol{\Gamma}$ is a diagonal matrix with diagonal elements {1, 0.5, 0.25, 0.125}.

```{r}
df %>% 
  filter(id %in% 1:4) %>% 
  mutate(probs = exp(eta_i)/(1+exp(eta_i))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_i), col = "blue")+
  geom_line(aes(x=sind, y=probs), col = "red")+
  geom_point(aes(x=sind, y=Y), size = 0.05)+
  facet_wrap(~id)+
  labs(title = "Four simulated subjects")
```

### Local GLMMS

The entire time grid into 100 small intervals with equal length. The cut is made on the order index of measurement time, but not the actually time itself. Each bin its labeled by its midpoint (5, 15, 25, ..., 995).

Within each interval labeled as $s$, we fit a local GLMM model:

$$logit(E(Y_i^s)) = \beta_0^s+b_i^s$$
Then we use this series of models to estimate the latent Gaussian function. Each subject will have a different value of latent function at each bin.

```{r}
source(here("Code/GLMM-FPCA.R"))
```

Figure below shows the estimated latent function for subjects 1-4:

```{r}
ggplot(df_pred_latent %>% filter(id %in% 1:4))+
  geom_line(aes(x = sind_inx, y=eta_hat))+
  geom_line(aes(x=sind_inx, y = eta_i, col = "red"))+
  # geom_point(aes(x=sind_inx, y = Y, col = "blue"), size = 0.01)+
  facet_wrap(~id)+
  labs(title = "Predicted latent functions on original grid")
```
```{r}
df_pred_latent %>% filter(id %in% 1:4 & sind_inx %in% mid) %>%
  ggplot()+
  geom_line(aes(x = sind_inx, y=eta_hat))+
  geom_line(aes(x=sind_inx, y = eta_i, col = "red"))+
  # geom_point(aes(x=sind_inx, y = Y, col = "blue"), size = 0.01)+
  facet_wrap(~id)+
  labs(title = "Predicted latent functions on binned grid")

@@

### FPCA

Here we fit FPCA on the estimated latent function from the previous step:

```{r}
## mean functions
plot(mid, fpca_fit$mu, ylab = "mean")
```

```{r, fig.height=8, fig.width=8}

## eigenfunctions
par(mfrow=c(2, 2))
plot(mid, fpca_fit$efunctions[, 1], ylab='PC1')
plot(mid, fpca_fit$efunctions[, 2], ylab='PC2')
plot(mid, fpca_fit$efunctions[, 3], ylab='PC3')
plot(mid, fpca_fit$efunctions[, 4], ylab='PC4')
```

### In-sample prediction

Now let's try to predict future outcomes based on observations $\boldsymbol{Y}$ up to $t_m$.

For in-sample prediction, we have observed the full outcome track, thus have estimated the full latent Gaussian function $\eta(s)$ on binned grid. 
For now, we will take the quantities up to $t_m$ and re-estimate the scores with empirical Bayes:

$$\hat{\boldsymbol{\xi}} = E(\boldsymbol{\xi}|\boldsymbol{Y}) =  \boldsymbol{\hat{\Gamma}\Phi ^T}(\boldsymbol{\Phi\hat{\Gamma}\Phi^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}(\hat{\boldsymbol{\eta}}-\hat{\boldsymbol{f}}_0)$$
Where:

- $\hat{\boldsymbol{\Gamma}}$ is the diagonal covariance matrix of eigenvalues
- $\boldsymbol{\Phi}$ is the matrix of eigenfunctions up to $t_m$. These should be constant across all subjects.
- $\hat{\sigma_{\epsilon}}^2$ is the residual variance from FPCA
- $\hat{\boldsymbol{\eta}}$ is the estimated latent Gaussian function up to $t_m$
- $\hat{\boldsymbol{f}}_0$ is the FPCA mean function up to $t_m$

Below we try to do that with one subject:

- Subject 1

```{r}
source(here("code/DynPred.R"))
```

```{r}
# function to clean results
# make predictions with partial track up to 20, 40, 60
clean_pred <- function(subj = 1){
  pred20 <- data.frame(bin = mid, 
                       eta_i = in_samp_dyn_pred(fpca_fit, 20,
                                                df_pred_unique, subj),
                       type = "up to 195") %>% 
    filter(bin>195)

  pred40 <- data.frame(bin = mid, 
                      eta_i = in_samp_dyn_pred(fpca_fit, 40,
                                               df_pred_unique, subj),
                       type = "up to 395") %>%
    filter(bin>395)

  pred60 <- data.frame(bin = mid, 
                       eta_i = in_samp_dyn_pred(fpca_fit, 60,
                                                df_pred_unique, subj),
                        type = "up to 595") %>%
    filter(bin>595)

  df_pred <- df_pred_latent %>% filter(id==subj & sind_inx==bin) %>%
    select(bin, eta_i) %>% mutate(type="true") %>%
    add_row(pred20) %>%
    add_row(pred40) %>% 
    add_row(pred60)

  return(df_pred)
}

```{r}
# Try on subjectw 1-4

clean_pred(1) %>%
  ggplot()+
  geom_line(aes(x=bin, y = eta_i, col=type))

clean_pred(2) %>%
  ggplot()+
  geom_line(aes(x=bin, y = eta_i, col=type))

clean_pred(3) %>%
  ggplot()+
  geom_line(aes(x=bin, y = eta_i, col=type))

clean_pred(4) %>%
  ggplot()+
  geom_line(aes(x=bin, y = eta_i, col=type))

```
- Subject 2

```{r}
# Try on subject 1
pred20 <- data.frame(bin = mid, 
                     eta_i = in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj=2),
                     type = "up to 195") %>% 
  filter(bin>195)

pred40 <- data.frame(bin = mid, 
                     eta_i = in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj=2),
                     type = "up to 395") %>%
  filter(bin>395)

pred60 <- data.frame(bin = mid, 
                     eta_i = in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj=2),
                     type = "up to 595") %>%
  filter(bin>595)

clean_pred(1)
  ggplot()+
  geom_line(aes(x=bin, y = eta_i, col=type))
```

- Subject 3

```{r}
# Try on subject 1
pred20 <- data.frame(bin = mid, 
                     eta_i = in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj=3),
                     type = "up to 195") %>% 
  filter(bin>195)

pred40 <- data.frame(bin = mid, 
                     eta_i = in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj=3),
                     type = "up to 395") %>%
  filter(bin>395)

pred60 <- data.frame(bin = mid, 
                     eta_i = in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj=3),
                     type = "up to 595") %>%
  filter(bin>595)

df_pred_latent %>% filter(id==3 & sind_inx==bin) %>%
  select(bin, eta_i) %>% mutate(type="true") %>%
  add_row(pred20) %>%
  add_row(pred40) %>% 
  add_row(pred60) %>% 
  ggplot()+
  geom_line(aes(x=bin, y = eta_i, col=type))
```
- Subject 4

```{r}
# Try on subject 1
pred20 <- data.frame(bin = mid, 
                     eta_i = in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj=4),
                     type = "up to 195") %>% 
  filter(bin>195)

pred40 <- data.frame(bin = mid, 
                     eta_i = in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj=4),
                     type = "up to 395") %>%
  filter(bin>395)

pred60 <- data.frame(bin = mid, 
                     eta_i = in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj=4),
                     type = "up to 595") %>%
  filter(bin>595)

df_pred_latent %>% filter(id==4 & sind_inx==bin) %>%
  select(bin, eta_i) %>% mutate(type="true") %>%
  add_row(pred20) %>%
  add_row(pred40) %>% 
  add_row(pred60) %>% 
  ggplot()+
  geom_line(aes(x=bin, y = eta_i, col=type))

@@

1. Interval prediction
2. FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?
3. Out-of-sample prediction: how to estimate scores?

Thoughts on 3:

1. If we'd like to make predictions on more than one sample, we may be able to fit GLMM on test samples and predict $\eta(t)$
2. Do a different set of derivation with MLE+EB like [this paper](https://www.jstor.org/stable/pdf/2533715.pdf?refreqid=excelsior%3A135dd765c41e4b46eb39bb6ef7a71053&ab_segments=&origin=&acceptTC=1)
